\thornado-hydro: Generalizing discontinuous Galerkin methods for a nuclear equation of state for supernova hydrodynamics

### 2019-05-06 paper/ms.tex
Submitted version

\noindent Where $H=(E+P)\tau$ is the specific enthalpy and
\beq
%  \begin{align}
    P_{\epsilon}  = \left(\frac{\partial{P}}{\partial{\epsilon}}\right)_{\tau,D_{e}}, \quad
    P_{D_e}  = \left(\frac{\partial{P}}{\partial{D_{e}}}\right)_{\tau, \epsilon}, \quad
    P_{\tau}  = \left(\frac{\partial{P}}{\partial{\tau}}\right)_{\epsilon, D_{e}}.
%  \end{align}
\eeq
The eigenvalues of the flux Jacobian are given by the diagonal matrix

@@

\begin{align*}
  \mathcal{R}_{1}^{-1} = \frac{1}{c^2}
  \left[
  \begin{array}{cccccc}
   \frac{1}{4} (2 c  v_{1}+\omega ) & \frac{1}{2} (-c- \phi_{1} ) & -\frac{1}{2} \phi_{2}
    & -\frac{1}{2} \phi_{3}  & \frac{P_{\epsilon} \tau }{2} & \frac{P_{D_{e}}}{2}
     \\
   -\frac{v_{2} \omega }{2} & \phi_{1} v_{2}  & c^2+\phi_{2} v_{2}  &
     \phi_{3} v_{2}  & -\phi_{2}  & -P_{D_{e}} v_{2}
     \\
   \frac{2 \chi  c^2+\alpha  \omega \tau^{-1} }{2 \chi } & -\frac{\phi_{1} \alpha  }{\chi \tau } &
     -\frac{\phi_{2} \alpha  }{\chi \tau } & -\frac{\phi_{3} \alpha }{\chi \tau } &
     \frac{P_{\epsilon} \alpha }{\chi } & \frac{P_{D_{e}} \left(\alpha -2 c^2\right)}{\tau \chi }
      \\
   -\frac{y_{e} P_{D_{e}} \omega }{\chi \tau } & \frac{2 y_{e} P_{D_{e}} \phi_{1} }{\chi \tau } & \frac{2 y_{e} P_{D_{e}}
     \phi_{2} }{\chi \tau} & \frac{2 y_{e} P_{D_{e}} \phi_{3} }{\chi \tau} & -\frac{2 y_{e}
     P_{D_{e}} P_{\epsilon} }{\chi } & \frac{2 P_{D_{e}} \left(c^2-y_{e} P_{D_{e}} \right)}{\tau  \chi }
      \\
   -\frac{v_{3} \omega }{2} & \phi_{1} v_{3}  & \phi_{2} v_{3}   &
     c^2+\phi_{3} v_{3}  & -\phi_{3}  & -P_{D_{e}} v_{3}
      \\
   \frac{1}{4} (\omega -2 c  v_{1}) & \frac{1}{2} (c-\phi_{1} ) & -\frac{1}{2} \phi_{2}
       & -\frac{1}{2} \phi_{3}  & \frac{P_{\epsilon} \tau }{2} & \frac{P_{D_{e}}}{2}
     \\
  \end{array}
  \right]
\end{align*}
where $\phi_{i} = P_{\epsilon}\,\tau\, v_{i}$,
$\omega = \tau\, (P_{\epsilon}\,(v^2 - 2\epsilon) - 2\,P_{\tau}\,\tau)$, and
$\alpha = 2 y_{e} P_{D_{e}} - \tau \chi$.
The left and right eigenvectors presented here are similar in form to those
presented in \citet{schaal:2015a}, but are considerably more complicated due to the
restriction of a nuclear EOS.

\section{Preliminary Numerical Results}
\label{sec:results}
In this section we present preliminary numerical results obtained with the DG
method implemented in \thornado\, leveraging use of the characteristic
decomposition. Unless otherwise noted, we use the SFHo EOS of
nuclear matter \citep{steiner:2013} now commonly used in high-fidelity CCSN simulations.
These tests serve to gauge the performance improvement of the
DG characteristic decomposition implementation on a set of benchmarks as an initial
assessment of its suitability for future CCSN simulations.

\subsection{Sod Shock Tube}
We present the classic 1D Sod shock tube Riemann problem \citep{sod:1978} in
Cartesian coordinates. This is an ideal first test as it is a fairly simple
scenario but contains both a contact discontinuity and a shock to test the
efficiency of our limiting scheme. The computational domain is D = [-5,5] km
with a discontinuity initially located at $x = 0$ km separating the left and right states

\begin{align*}
  \mathbf{U}_{L} &= (10^{12}~\text{g~cm}^{-3}, 0\,, 3.712*10^{32}~\text{ergs~cm}^{-3}, 0.4*10^{12})^T\,\,\, \\
  \mathbf{U}_{R} &= (1.25*10^{11}~\text{g~cm}^{-3}, 0\, , 3.015*10^{31}~\text{ergs~cm}^{-3}, 0.375*10^{11})^T.
\end{align*}
\footnote{\textbf{Fix e in the state vectors}}
\noindent The test is run until $t = 0.025$ms with 100 elements
using $C_{\TCI} = 0.2$ and $\beta_{\TVD} = 2.0$. Results are plotted in
Figure~\ref{fig:SodSedovOptimal}. Using the characteristic limiting described
above, the DG method captures the features of the solution,
resolving well both the contact discontinuity and shock, without introducing
noticeable oscillations in the numerical solutions near the discontinuities.
In Figure~\ref{fig:shock}
we plot elements in the $xt$-plane flagged by the troubled cell indicator for
limiting, showing that the troubled cell indocator flagged cells around both
the shock and contact discontinuity for limiting and maintained that
for the full time of the test.

\begin{figure}[h!]
  \centering
  \includegraphics[width=32pc,height=30pc]{optimal.png}
  \centering
  \caption{\label{fig:SodSedovOptimal} Numerical solution of Sod's problem using
    100 elements compared with a reference solution using 10000 elements.}
\end{figure}
%\clearpage
\begin{wrapfigure}{L}{2.8in}
  \includegraphics[width=18pc]{optimal_shock.png}
  \caption{$xt$-plane of elements flagged for limiting by the troubled cell indicator.}
  \label{fig:shock}
\end{wrapfigure}

\subsection{Optimal Limiting Parameters}
\label{sec:param}
In order to determine the optimal limiting parameters $C_{\TCI}$ and $\beta_{\TVD}$,
we performed a suite of sixteen simulations of Sod's problem with various
limiting parameters with $\beta_{\TVD}\in\left[1.0,2.0\right]$ and
$C_{\TCI}\in\left[0.0,0.2\right]$. For each $\left(\beta_{\TVD},C_{\TCI}\right)$ pair,
we computed the relative error and total variation \textbf{\textcolor{red}{EQUATIONS}} in density and electron
fraction at $t = 0.025$ms. Results are plotted in Figure~\ref{fig:landscape}.
We find that, as expected, increasing $\beta_{\TVD}$ tends to monotonically
decrease the relative error while (non-monotonically) increasing the total variation due to the
limiter allowing for more oscillations. Increasing $C_{\TCI}$ tends to decrease
the relative error and increase the total variation.
We have selected
$\beta_{\TVD}=1.75$ and $C_{\TCI}=0.1$ to be the optimal parameters providing
the best combination of relative error and total variation reduction. Unless
otherwise noted, all following results will use this combination of limiting
parameters.

\begin{figure}[b]
  \centering
  \includegraphics[width=32pc, height=27.5pc]{./figures/error_tv_landscape.png}
  \caption{\label{fig:landscape} Error - total variation landscapes for various
  values of the limiting parameters $C_{TCI}$ and $\beta_{TVD}$.}
\end{figure}

\subsection{Improvement From Componentwise Limiting}
\label{sec:optimal}
The motivation for limiting on the characteristic variables is the potential
improvement from componentwise limiting. In Figure~\ref{fig:SodSedovCW} we plot the
solution at $t = 0.025$ms using 100 elements employing both characteristic
limiting (blue) and componentwise limiting (red) compared to a reference
solution computed with 10000 elements. We observe, for the componentwise limiting,
noticable oscillations in density around the contact discontinuity as well
as less resolved discontinuities. The electron fraction displays very
large oscillations at the left side of the contact discontinuity, up to
$\pm 0.02$ in the cell averaged electron fraction. This makes characteristic
limiting particularly appealing for CCSN simulations where the electron fraction
plays a critical role in the explosion dynamics.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=36pc]{./figures/characteristic_cw.png}
  \caption{\label{fig:SodSedovCW} Numerical solution of Sod's problem using
  100 elements with characteristic limiting (blue) and componentwise limiting
  (red) compared with a reference solution using 10000 elements.}
\end{figure}

\subsection{High Density Regime}
This test is included to verify the performance improvements of the characteristic
limiting in a higher density regime. It is designed with the Sod problem in mind.
The computational domain is $D = [-5,5]$ km
with a discontinuity initially located at $x = 0$ km separating the left and right states

\begin{align*}
  \mathbf{U}_{L} &= (10^{13}~\text{g~cm}^{-3}, 0\,, 3.712*10^{32}~\text{ergs~cm}^{-3}, 0.15*10^{12})^T\,\,\, \\
  \mathbf{U}_{R} &= (1.25*10^{12}~\text{g~cm}^{-3}, 0\, , 3.015*10^{31}~\text{ergs~cm}^{-3}, 0.169*10^{12})^T.
\end{align*}

\noindent The test is run until $t = 0.05$ms with 100 elements
using $C_{\TCI} = 0.1$ and $\beta_{\TVD} = 1.2$. We chose a lower $\beta_{\TVD}$
here, as larger values resulted in unphysical internal energies outside of the
EOS table causing the test to fail, motivating the development of a
positivity limiter compatible with a tabulated nuclear matter EOS.
Inital state values were chosen to be consistent with
high-fidelity CCSN simulation code \chimera\, \citep{bruenn:2018} data.
Results are plotted in
Figure~\ref{fig:SodSedovHighDens}. Our method captures the main features
of the solution, but introduces more oscillations than in the lower density
case, even when using the characteristic limiter.
We notice an undershooting of the right side
of the contact discontinuity in the electron fraction not previously present
warranting further investigation.

@@

\subsection{EOS Resolution Dependence}
\label{sec:eosRes}
In order to test the sensitivity of the limiter on the EOS table resolution,
we repeated the tests in Section~\ref{sec:optimal} with a higher resolution
SFHo EOS table. As before, tests are computed using 100 elements until
$t = 0.025$ms with characteristic limiting and the optimal limiting parameters determined in
Section~\ref{sec:param} and compared to a reference solution computed using
10000 elements. Results for both tables are plotted in Figure~\ref{fig:SodSedovSFHoRes}.
We find nearly no sensitivity to the table resolution, with the exception of
minor differences in density to the left of the contact discontinuity.
\begin{figure}[h!]
  \centering
  \includegraphics[width=36pc]{./figures/eos_res.png}
  \caption{\label{fig:SodSedovSFHoRes} Comparison of numerical solutions to
  Sod's problem using 100 elements with characteristic limiting and two
  different EOS table resolutions.}
\end{figure}

\subsection{EOS Table Sensitivity}
In order to probe the sensitivity of the method to the EOS table, we repeated
the tests conducted in Section~\ref{sec:optimal} using three different EOS tables:
the high resolution SFHo table from Section~\ref{sec:eosRes}, the SFHx table
(\textbf{citation}), and the DD2 table (\textbf{citation}).
As before, tests are computed using 100 elements until
$t = 0.025$ms with characteristic limiting and the optimal limiting parameters determined in
Section~\ref{sec:param}. We find nearly no sensitivity to the EOS table used in
the low density regime, with the exception of very small variations in density
to the left of the contact discontinuity. Future work includes testing the
sensitivity to the EOS table in higher density regimes.
\begin{figure}[h!]
  \centering
  \includegraphics[width=36pc]{./figures/eos_all.png}
  \caption{\label{fig:SodSedovEOS} Comparison of numerical solutions to
  Sod's problem using 100 elements with characteristic limiting and three
  different EOS tables.}
\end{figure}
%\FloatBarrier

@@

We have presented preliminary developments and numerical results for a
characteristic limiter to be used with solvers of the non-relativistic
Euler equations of gas dynamics in the {\bf t}oolkit for {\bf h}igh-{\bf or}der
{\bf n}eutrino-r{\bf ad}iation hydr{\bf o}dynamics (\thornado). The results
presented from a suite of 1D test problems, based around Sod's 1D shock
tube problem, demonstrate the performance of the characteristic limiter
compared to the componentwise limiter. Moreover, after performing a parameter
study on the limiter parameters, we found optimal limiting
parameters for use with the Sod problem. While the optimal limiting parameters
tend to be problem dependent, we hope to investigate if such an optimal
parameter exists for CCSN applications. Tests with various EOS tables and
table resolutions showed little to no sensitivity to the table or its
resolution in the tests studied. For future work, we will extend these studies
to other higher density regimes applicable to the CCSN environment.
Planned near-future work on the nuclear equation of state compatibility of
\thornado\, includes the development of a positivity limiter and extension of
the characteristic limtier to handle relativistic and curvilinear problems.

\acknowledgements
We thank people.

@@

\section{Derivatives}
\input{derivatives}

>>>>

\noindent Where $H=(E+P)\tau$ is the specific enthalpy and
\beq
%  \begin{align}
    P_{\epsilon}  = \left(\frac{\partial{P}}{\partial{\epsilon}}\right)_{\tau,D_{e}}, \quad
    P_{D_e}  = \left(\frac{\partial{P}}{\partial{D_{e}}}\right)_{\tau, \epsilon}, \quad
    P_{\tau}  = \left(\frac{\partial{P}}{\partial{\tau}}\right)_{\epsilon, D_{e}}
%  \end{align}
\eeq
are the necessary thermodynamic derivatives. Expressions for these derivatives
in terms of the table variables $\rho, T, \text{and}~y_{e}$ are given in Appendix~\ref{appendix:deriv}.
The eigenvalues of the flux Jacobian are given by the diagonal matrix

@@

\begin{align*}
  \mathcal{R}_{1}^{-1} = \frac{1}{c^2}
  \left[
  \begin{array}{cccccc}
   \frac{1}{4} (2 c  v_{1}+\omega ) & \frac{1}{2} (-c- \phi_{1} ) & -\frac{1}{2} \phi_{2}
    & -\frac{1}{2} \phi_{3}  & \frac{P_{\epsilon} \tau }{2} & \frac{P_{D_{e}}}{2}
     \\
   -\frac{v_{2} \omega }{2} & \phi_{1} v_{2}  & c^2+\phi_{2} v_{2}  &
     \phi_{3} v_{2}  & -\phi_{2}  & -P_{D_{e}} v_{2}
     \\
   \frac{2 \chi  c^2+\alpha  \omega \tau^{-1} }{2 \chi } & -\frac{\phi_{1} \alpha  }{\chi \tau } &
     -\frac{\phi_{2} \alpha  }{\chi \tau } & -\frac{\phi_{3} \alpha }{\chi \tau } &
     \frac{P_{\epsilon} \alpha }{\chi } & \frac{P_{D_{e}} \left(\alpha -2 c^2\right)}{\tau \chi }
      \\
   -\frac{y_{e} P_{D_{e}} \omega }{\chi \tau } & \frac{2 y_{e} P_{D_{e}} \phi_{1} }{\chi \tau } & \frac{2 y_{e} P_{D_{e}}
     \phi_{2} }{\chi \tau} & \frac{2 y_{e} P_{D_{e}} \phi_{3} }{\chi \tau} & -\frac{2 y_{e}
     P_{D_{e}} P_{\epsilon} }{\chi } & \frac{2 P_{D_{e}} \left(c^2-y_{e} P_{D_{e}} \right)}{\tau  \chi }
      \\
   -\frac{v_{3} \omega }{2} & \phi_{1} v_{3}  & \phi_{2} v_{3}   &
     c^2+\phi_{3} v_{3}  & -\phi_{3}  & -P_{D_{e}} v_{3}
      \\
   \frac{1}{4} (\omega -2 c  v_{1}) & \frac{1}{2} (c-\phi_{1} ) & -\frac{1}{2} \phi_{2}
       & -\frac{1}{2} \phi_{3}  & \frac{P_{\epsilon} \tau }{2} & \frac{P_{D_{e}}}{2}
     \\
  \end{array}
  \right]
\end{align*}
where $\phi_{i} = P_{\epsilon}\,\tau\, v_{i}$,
$\omega = \tau\, (P_{\epsilon}\,(v^2 - 2\epsilon) - 2\,P_{\tau}\,\tau)$, and
$\alpha = 2 y_{e} P_{D_{e}} - \tau \chi$.
The left and right eigenvectors presented here are similar in form to those
presented in \citet{schaal:2015a}, but are considerably more complicated due to the
restriction of a nuclear EOS. Having an explicit form for these matrices is
desirable as it eliminates the need to numerically diagonalize the flux
Jacobian\footnote{These matrices and their derivations are available in Mathematica
notebooks at \url{https://github.com/AstroBarker/dgHydro-nuc_charDecomp}}.

\section{Preliminary Numerical Results}
\label{sec:results}
In this section we present preliminary numerical results obtained with the DG
method implemented in \thornado\, leveraging use of the characteristic
decomposition. Unless otherwise noted, we use the SFHo EOS of
nuclear matter \citep{steiner:2013} now commonly used in high-fidelity CCSN simulations
due to its consistency with neutron star mass and radius observations.
These tests serve to gauge the performance improvement of the
DG characteristic decomposition implementation on a set of benchmarks as an initial
assessment of its suitability for future CCSN simulations.

\subsection{Riemann Problem}
\label{sec:results:1}
We present a 1D Riemann problem inspired by the classic 1D Sod shock tube Riemann problem \citep{sod:1978} in
Cartesian coordinates. This is an ideal first test as it is a fairly simple
scenario but contains both a contact discontinuity and a shock to test the
efficiency of our limiting scheme, though due to the nature of our EOS, an analytic
solution does not exist. The computational domain is D = [-5,5]~km
with a discontinuity initially located at $x = 0$~km separating the left and right states

\begin{align*}
  \mathbf{U}_{L} &= (10^{12}~\text{g~cm}^{-3}, 0\,, 2.703\text{x}10^{32}~\text{ergs~cm}^{-3}, 0.4\text{x}10^{12})^T\,\,\, \\
  \mathbf{U}_{R} &= (1.25\text{x}10^{11}~\text{g~cm}^{-3}, 0\, , 2.822\text{x}10^{31}~\text{ergs~cm}^{-3}, 0.375\text{x}10^{11})^T.
\end{align*}

\noindent The test is run until $t = 0.025$~ms with 100 elements
using $C_{\TCI} = 0.2$ and $\beta_{\TVD} = 2.0$ with a third-order time
integration SSP-RK3 scheme. Results for density, pressure, velocity, and electron fraction
are plotted in the upper left, upper right, lower left, and lower right panels of
Figure~\ref{fig:SodSedovOptimal}, respectively, compared to a reference run using
10000 elements, third-order time integration, and first order spacial discretization.
Using the characteristic limiting described
above, the DG method captures the features of the solution,
resolving well the contact discontinuity, shock, and rarefraction wave
located at about $x=2$~km, $x=4$~km, and from $x=-3$~km to $x=0$~km, respectively, without introducing
noticeable oscillations in the numerical solutions near the discontinuities.
In Figure~\ref{fig:shock}
we plot elements in the $xt$-plane flagged by the troubled cell indicator for
limiting, showing that the troubled cell indocator flagged cells around both
the shock and contact discontinuity for limiting and maintained that
for the full time of the test.

\begin{figure}[h!]
  \centering
  \includegraphics[width=32pc,height=30pc]{optimal.png}
  \centering
  \caption{\label{fig:SodSedovOptimal} Numerical solution of the Riemann problem using
    100 elements and $C_{\TCI} = 0.2$ and $\beta_{\TVD} = 2.0$ with a third-order
    time integration SSP-RK3 scheme for density (upper left), pressure (upper right),
    velocity (lower left), and electron fraction (lower right)
    compared with a reference solution (black) using 10000 elements.}
\end{figure}
%\clearpage
\begin{wrapfigure}{L}{2.8in}
  \includegraphics[width=18pc]{optimal_shock.png}
  \caption{$xt$-plane of elements flagged for limiting by the troubled cell indicator
  in the Riemann problem of Section~\ref{sec:results:1}.}
  \label{fig:shock}
\end{wrapfigure}

\subsection{Optimal Limiting Parameters}
\label{sec:param}
In order to determine the optimal limiting parameters $C_{\TCI}$ and $\beta_{\TVD}$,
we performed a suite of sixteen simulations of Sod's problem with various
limiting parameters with $\beta_{\TVD}\in\left[1.0,2.0\right]$ and
$C_{\TCI}\in\left[0.0,0.2\right]$. For each $\left(\beta_{\TVD},C_{\TCI}\right)$ pair,
we computed the relative error and total variation \
\beq
  \varepsilon(\mathbf{U}) = \frac{1}{N}\sum_{\bK}\left\lvert\frac{\vect{U}_{h}^{\text{ref}}(x_{q}) - \vect{U}_{h}(x_{q})}{\vect{U}_{h}^{\text{ref}}(x_{q})}\right\rvert \quad\quad
  TV = \sum_{k}\abs{ \bar{\vect{U}}_{k+1} - \bar{\vect{U}}_{k} }
\eeq
 in density and electron
fraction at $t = 0.025$ms, where $\vect{U}_{h}^{\text{ref}} $ is the reference
solution, $\bar{\vect{U}}_{k+1}$ is the cell average is cell $k$, and $x_{q}$ are
numerical quadrature points.
Results are plotted in Figure~\ref{fig:landscape}.
We find that, as expected, increasing $\beta_{\TVD}$ tends to monotonically
decrease the relative error while (non-monotonically) increasing the total variation due to the
limiter allowing for more oscillations. Increasing $C_{\TCI}$ tends to decrease
the relative error and increase the total variation.
We have selected
$\beta_{\TVD}=1.75$ and $C_{\TCI}=0.1$ to be the optimal parameters providing
the best combination of relative error and total variation reduction. Unless
otherwise noted, all following results will use this combination of limiting
parameters.

\begin{figure}[b]
  \centering
  \includegraphics[width=32pc, height=27.5pc]{./figures/error_tv_landscape.png}
  \caption{\label{fig:landscape} Error and total variation landscapes for various
  values of the limiting parameters $C_{TCI}$ and $\beta_{TVD}$. The top row shows
  total variation in density (left) and electron fraction (right). The bottom
  row show relative error in density (left) and electron fraction (right).}
\end{figure}

\subsection{Improvement From Componentwise Limiting}
\label{sec:optimal}
The motivation for limiting on the characteristic variables is the potential
improvement from componentwise limiting. In Figure~\ref{fig:SodSedovCW} we plot the
solution at $t = 0.025$~ms using 100 elements employing both characteristic
limiting (blue) and componentwise limiting (red) compared to the reference
solution computed with 10000 elements. We observe, for the componentwise limiting,
noticable oscillations in density around the contact discontinuity as well
as less resolved discontinuities. The electron fraction displays very
large oscillations at the left side of the contact discontinuity, up to
$\pm 0.02$ in the cell averaged electron fraction. The solutions with
characteristic limiting are very close to the reference solution, presenting
a significant improvement over componentwise limiting.
This makes characteristic
limiting particularly appealing for CCSN simulations where the electron fraction
plays a critical role in the explosion dynamics.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=36pc]{./figures/characteristic_cw.png}
  \caption{\label{fig:SodSedovCW} Numerical solution of Sod's problem using
  100 elements with characteristic limiting (blue) and componentwise limiting
  (red) for density (upper left), pressure (upper right),
  velocity (lower left), and electron fraction (lower right)
  compared to a reference solution (black) using 10000 elements.compared with a reference solution using 10000 elements.}
\end{figure}

\vspace{1cm}
\subsection{High Density Regime}
This test is included to verify the performance improvements of characteristic
limiting in a higher density regime. It is designed with the Sod problem in mind.
The computational domain is $D = [-5,5]$~km
with a discontinuity initially located at $x = 0$~km separating the left and right states

\begin{align*}
  \mathbf{U}_{L} &= (10^{13}~\text{g~cm}^{-3}, 0\,, 3.712\text{x}10^{32}~\text{ergs~cm}^{-3}, 0.15\text{x}10^{12})^T\,\,\, \\
  \mathbf{U}_{R} &= (1.25\text{x}10^{12}~\text{g~cm}^{-3}, 0\, , 3.015\text{x}10^{31}~\text{ergs~cm}^{-3}, 0.169\text{x}10^{12})^T.
\end{align*}

\noindent The test is run until $t = 0.05$~ms with 100 elements
using $C_{\TCI} = 0.1$ and $\beta_{\TVD} = 1.2$. We chose a lower $\beta_{\TVD}$
here, as larger values resulted in unphysical internal energies outside of the
EOS table causing the test to fail, motivating the development of a
positivity limiter compatible with a tabulated nuclear matter EOS \citep{zhang:2010a}.
Inital state values were chosen to be consistent with
high-fidelity CCSN simulation code \chimera\, \citep{bruenn:2018} data by first
choosing the left and right densities to be an order of magnitude larger than in
Section~\ref{sec:results:1}, and then finding temperatures and electron fractions
consistent with those densities from \chimera\, data, and computing the other
state variables from the EOS table.
Results are plotted in
Figure~\ref{fig:SodSedovHighDens}. Our method captures the main features
of the solution, but introduces more oscillations than in the lower density
case, even when using the characteristic limiter.
We notice an undershooting of the right side
of the contact discontinuity in the electron fraction not previously present
warranting further investigation.

@@

\subsection{EOS Resolution Dependence}
\label{sec:eosRes}
In order to test the sensitivity of the limiter on the EOS table resolution,
we repeated the tests in Section~\ref{sec:optimal} with a higher resolution
SFHo EOS table. As before, tests are computed using 100 elements until
$t = 0.025$ms with characteristic limiting and the optimal limiting parameters determined in
Section~\ref{sec:param} and compared to a reference solution computed using
10000 elements. Results for both tables are plotted in Figure~\ref{fig:SodSedovSFHoRes}.
We find no sensitivity to the table resolution, with the exception of
minor differences in density to the left of the contact discontinuity.
\begin{figure}[h!]
  \centering
  \includegraphics[width=36pc]{./figures/eos_res.png}
  \caption{\label{fig:SodSedovSFHoRes} Comparison of numerical solutions to
  Sod's problem using 100 elements with characteristic limiting and two
  different EOS table resolutions (low: red, high: black) for density (upper left), pressure (upper right),
  velocity (lower left), and electron fraction (lower right)
  compared with a reference solution (black) using 10000 elements and the low resolution EOS table.}
\end{figure}

\subsection{EOS Table Sensitivity}
In order to probe the sensitivity of the method to the EOS table, we repeated
the tests conducted in Section~\ref{sec:optimal} using three different EOS tables:
the high resolution SFHo table from Section~\ref{sec:eosRes}, the SFHx table
(\textbf{citation}), and the DD2 table (\textbf{citation}).
As before, tests are computed using 100 elements until
$t = 0.025$ms with characteristic limiting and the optimal limiting parameters determined in
Section~\ref{sec:param}. We find nearly no sensitivity to the EOS table used in
the low density regime, with the exception of very small variations in density
to the left of the contact discontinuity. Future work includes testing the
sensitivity to the EOS table in higher density regimes.
\begin{figure}[h!]
  \centering
  \includegraphics[width=36pc]{./figures/eos_all.png}
  \caption{\label{fig:SodSedovEOS} Comparison of numerical solutions to
  Sod's problem using 100 elements with characteristic limiting and three
  different EOS tables for density (upper left), pressure (upper right),
  velocity (lower left), and electron fraction (lower right).
  }
\end{figure}
%\FloatBarrier

@@

We have presented preliminary developments and numerical results for a
characteristic limiter to be used with solvers of the non-relativistic
Euler equations of gas dynamics in the {\bf t}oolkit for {\bf h}igh-{\bf or}der
{\bf n}eutrino-r{\bf ad}iation hydr{\bf o}dynamics (\thornado). We presented analytic
forms for the diagonalizating matrices for the flux Jacobian of the Euler equations
consistent with a tabulated nuclear EOS.
The results presented from a suite of 1D test problems
demonstrate the superior performance of the characteristic limiter
compared to the componentwise limiter. Moreover, after performing a parameter
study on the limiter parameters, we found optimal limiting
parameters for use with the Sod problem. While the optimal limiting parameters
tend to be problem dependent, we hope to investigate if such an optimal
parameter exists for CCSN applications. Tests with various EOS tables and
table resolutions showed little to no sensitivity to the table or its
resolution in the tests studied. For future work, we will extend these studies
to other higher density regimes applicable to the CCSN environment.
Planned near-future work on the nuclear equation of state compatibility of
\thornado\, includes the development of a positivity limiter and extension of
the characteristic limtier to handle multi-dimensional, relativistic, and curvilinear problems.
Because of the superior performance of characteristic limiting, we hope that
this work will improve the fidelity of CCSN simulations.

\acknowledgments{
Eirik Endeve and Anthony Mezzacappa acknowledge support from the NSF Gravitational Physics Program (NSF-GP 1505933 and 1806692).
}

@@

\section{Derivatives}
\label{appendix:deriv}
\input{derivatives}
